---
title: "SQL 101 For Everyone!"
author: "Simon L"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
#install.packages('DBI')
#install.packages('RSQLite')
library(RSQLite)
library(DBI)
```

## A quick, easy walkthrough for SQL!

SQL is a worldwide language used to manage data that is held in relational databases. The good thing is that we can handle SQL databases in R studio. To achieve this, we need to make sure we have the required packages

Once we have our package installed, we are going to make a connection to an SQL Server.

There are some things you need to keep in mind, let's clarify:

-   You need a sql database, make sure you known the information of the different databases you are going to use.

-   In case you want to create your own database and add files to your database follow the following short tutorial

### Making a database and adding files

1.  Create a connection and a database. In this example I am creating a database called `genomic_database.sqlite` in the Documents folder.

Now there's an open connection

```{r, eval = F}
connect = dbConnect(RSQLite::SQLite(), '~/Documents/genomic_database.sqlite')
```

2.  Adding files. Choose whichever file you want. Once you have the file in the format you want, you can save it in your database by using the function `dbWriteTable` and assign the names you prefer. In this example I am adding two dataframes, one that has information about downsizing factors based on a previously established optimal sequencing depth and another one that has information about files names and experiments.

```{r, eval = F}
data = read.delim('file.txt', sep = ',')
dbWriteTable(connect, 'downsize_factors', data, overwrite = T)

data = read.delim('file2.txt', sep = '\t')
dbWriteTable(connect, 'dataframe_experiment_info', data, overwrite = T)
```

This database can be found on the following directory on github <https://github.com/slizarazo21/didactic_code/tree/main/r_markdown>

### Establishing a connection and making SQL queries. 

Now that we have our database we are ready to start our analysis. Let's start a connection specifying what database we want to work with

```{r}
connect = dbConnect(SQLite(), '~/Documents/genomic_database.sqlite')

```

#### A SELECT Queries.

The basic structure for a select query on sql goes as follows

```{sql, connection = connect, eval = F}

SELECT column1, column2, ...
FROM mydataframe
WHERE a condition for your analysis
ORDER BY column1, column2 ASC/DESC
```

Our querie is very intuitive, we are subseting a specific column (or columns) from a data frame and we want to see the data that exists based on a condition.

Let's start with observing all the columns present in the dataframe downsize factors. For that the token `*` will select everything!

```{sql, connection = connect}

SELECT *
FROM downsize_factors
```

Now let's select only the column Experiment Seqdepth. Pretty straight-forward right?.

```{sql, connection = connect}
SELECT Experiment, Seqdepth
FROM downsize_factors
```

Let's include some conditions. The command `WHERE` requires different operators, a list of some useful operators can be found here <https://www.w3schools.com/sql/sql_where.asp> (With more examples too!).

Let's find the observations where the `factor` column is less than 0.5 and let's display only the columns `Experiment` and `Seqdepth`

```{sql, connection = connect}
SELECT Experiment, Seqdepth
FROM downsize_factors
WHERE factor < 0.5
```

We can also use other operators like `BETWEEN`. Let's look for those that are between 0.55 and 0.56

```{sql, connection = connect}
SELECT Experiment, Seqdepth, factor
FROM downsize_factors
WHERE factor BETWEEN 0.55 AND 0.56
```

We can start making more complex queries. For example let's order the previous query by `Seqdepth`, for that you can use `ORDER BY` and specify the ordering with `ASC/DESC`

```{sql, connection = connect}
SELECT Experiment, Seqdepth, factor
FROM downsize_factors
WHERE factor BETWEEN 0.55 AND 0.56
ORDER BY Seqdepth DESC
```

Moreover we can look for string matching. In that dataframe we have different types of cancers. Let's make a query where we are filtering by `factor` \> 0.5, but we want to look for BRCA cancer. For that we can use `AND/OR` and other string matching attributes like `LIKE`, `%%` and `_` **(These last attributes ALWAYS need to have the `LIKE` command)**

The %% operator, will look for matching strings. For example, vector A has the following strings

`A = c( 'CASA', 'MASA', 'LAZO', 'TAZO', 'AMASAR')`

If in my querie I put `WHERE A LIKE '%AS%'` I will get all the words that match the string `AS`. Thus we would get `CASA, MASA, AMASAR`

```{sql, connection = connect}
SELECT Experiment, Seqdepth, factor
FROM downsize_factors
WHERE factor > 0.5
  AND experiment LIKE '%BRCA%'
ORDER BY Seqdepth DESC
```

Let's subset for two specific types of cancer, for example COAD and LGG. In this case our condition query has the following form:

`( condition 1 AND (condition 2))`

`Where condition 2 = (Condition 3 OR condition 4)`

```{sql, connection = connect}
SELECT Experiment, Seqdepth, factor
FROM downsize_factors
WHERE factor > 0.5
  AND (experiment LIKE '%COAD%' OR experiment LIKE '%LGG%')
ORDER BY Seqdepth DESC
```

```{r}
dbDisconnect(connect)
```

